Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-01-11 ms/manuscript.Rmd
Use latex (mostly via gt) for all tables

# Read in csv version of the table
traits_used <- read_csv(here::here("data/table_1.csv"))

# Write out table in html for submitting
# MS as word doc
traits_used %>%
  gt(rowname_col = "trait") %>%
  fmt_markdown(columns = everything()) %>%
  tab_row_group(
    group = "Sporophyte traits",
    rows = matches("Stipe|Frond|Rhizome|Pinna|Leaf")
  ) %>%
  tab_row_group(
    group = "Gametophyte traits",
    rows = matches("Morphotype|Gemmae|Glands|Hairs")
  ) %>%
  cols_label(
    data_type = "Data type (unit)",
    ecological_significance = "Ecological Significance",
    reference = "Reference(s)"
  ) %>% 
  gtsave(here::here("ms/Table_1.html"))

# Include in PDF using kableExtra to convert to latex format
traits_used %>% 
  rename(`Data type (unit)` = data_type, `Functional significance` = ecological_significance, `Reference(s)` = reference) %>%
  column_to_rownames("trait") %>%
  knitr::kable(caption = "", format = "latex", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows("Sporophyte traits", 1, 7) %>%
  pack_rows("Gametophyte traits", 8, 11) %>%
  column_spec(2, width = "8em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em") %>%
  footnote(general = "Noncordate gametophyte morphotypes include ribbon, strap and filamentous.", general_title = "", threeparttable = TRUE, escape = TRUE)
```

@@

# Format the raw results for table presentation
phylosig_cont_traits_table <-
phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait)

# Write out table in html format
phylosig_cont_traits_table %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    kval = md("*K*"),
    k_pval = md("*P*(*K*)"),
    lambda_pval = md("*P*(lambda)"),
    lambda = "lambda"
  )  %>%
  gtsave(here::here("ms/Table_2.html"))

# Include in PDF in latex format
phylosig_cont_traits_table %>%
  rename(K = kval, `P(K)` = k_pval, `P(lambda)` = lambda_pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "", format = "latex", booktabs = TRUE) %>%
  footnote(general = "K, Blomberg's \\\\textit{K} (Blomberg \\\\textit{et al.}, 2003); lambda, Pagel's lambda (Pagel, 1999). Values of lambda or \\\\textit{K} greater than 1 indicate traits evolving according to Brownian Motion (BM); values of lambda or \\\\textit{K} near 0 indicate random distribution of traits.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

@@

# Format the raw results for table presentation
phylosig_binary_traits_table <-
phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait)

# Write out table in html
phylosig_binary_traits_table %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    num_present = "Number of presences",
    num_absent = "Number of absences",
    D = md("*D*"),
    prob_random = md("*P*(rnd)"),
    prob_brownian = md("*P*(bm)")
  ) %>% 
  gtsave(here::here("ms/Table_3.html"))

# Include in PDF in latex format
phylosig_binary_traits_table %>%
  select(trait, `Number of presences` = num_present, `Number of absences` = num_absent,
         D, `P(rnd)` = prob_random, `P(bm)` = prob_brownian) %>%
column_to_rownames("trait") %>%
  kable(caption = "", format = "latex", booktabs = TRUE) %>%
  footnote(general = "D, Fritz and Purvis’ \\\\textit{D} (Fritz \\\\& Purvis, 2010); P(rnd), probability of random distribution of traits; P(bm), probability of traits evolving under Brownian motion (BM). Values of \\\\textit{D} 0 or less indicate evolution according to BM; values near 1 indicate random trait distribution.", general_title = "", threeparttable = TRUE, escape = FALSE)
```

@@

# Format the raw results for table presentation
pic_table <- 
pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait)

# Write out table in html
pic_table %>%
gt(rowname_col = "trait") %>%
  cols_label(
    n_contrasts = "Num. constrasts",
    n_pos_con = "Num. pos. constrasts",
    statistic = md("*t*"),
    p.value = md("*P*"),
  ) %>% 
  gtsave(here::here("ms/Table_4.html"))

# Include in PDF in latex format
pic_table %>%
  rename(t = statistic, P = p.value, `Num. constrasts` = n_contrasts, `Num. pos. constrasts` = n_pos_con) %>%
  column_to_rownames("trait") %>%
  kable(caption = "", format = "latex", booktabs = TRUE) %>%
  footnote(general = "All trait contrasts were made by subtracting epiphytic values from terrestrial values; thus, a positive contrast indicates greater values for terrestrial clades.", general_title = "", threeparttable = TRUE)
```

@@

# Format the raw results for table presentation
correlated_evo_test_table <-
correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait)

# Write out table in rich text format
correlated_evo_test_table %>%
gt(rowname_col = "trait") %>%
  cols_label(
    logL_indep = "LogL(independent)",
    logL_dep = "LogL(dependent)",
    likelihood_ratio = "Likelihood ratio",
    pval = md("*P*"),
  ) %>% 
  gtsave(here::here("ms/Table_5.html"))

# Include in PDF in latex format
correlated_evo_test_table %>%
  rename(`LogL(independent)` = logL_indep, `LogL(dependent)` = logL_dep, `Likelihood ratio` = likelihood_ratio, P = pval) %>%
  column_to_rownames("trait") %>%
  kable(caption = "", format = "latex", booktabs = TRUE) %>%
  footnote(general = "LogL, log-likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial growth). Traits with a significantly lower log likelihood for the dependent model are significantly correlated with growth habit.", general_title = "", threeparttable = TRUE)
```

>>>>

# Use kable instead of gt because gt can't control table width,
# and the "ecological significance" descriptions would run off the page in PDF.

# FIXME: need to manually add "Sporophyte traits" and "Gametophyte traits" to
# table in docx (this happens with gt or kable).
read_csv(here::here("data/table_1.csv")) %>%
  rename(`Data type (unit)` = data_type, `Functional significance` = ecological_significance, `Reference(s)` = reference) %>%
  column_to_rownames("trait") %>%
  knitr::kable(format = "latex", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows("Sporophyte traits", 1, 7) %>%
  pack_rows("Gametophyte traits", 8, 11) %>%
  column_spec(2, width = "8em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em")
```

Noncordate gametophyte morphotypes include ribbon, strap and filamentous.

@@

phylosig_cont_traits %>%
  mutate_at(vars(kval, lambda), ~round_t(., 2)) %>%
  mutate(lambda_pval = format_pval(lambda_pval)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait) %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    kval = md("*K*"),
    k_pval = md("*P*(*K*)"),
    lambda_pval = md("*P*(lambda)"),
    lambda = "lambda"
  )  %>%
  as_latex

\K, Blomberg's *K* [@Blomberg2003]; lambda, Pagel's lambda [@Pagel1999]. Values of lambda or \K greater than 1 indicate traits evolving according to Brownian Motion (BM); values of lambda or \K near 0 indicate random distribution of traits.

@@

phylosig_binary_traits %>%
  mutate(D = round_t(D, 2)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("habit", "gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Epiphytic growth", "Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    num_present = "Number of presences",
    num_absent = "Number of absences",
    D = md("*D*"),
    prob_random = md("*P*(rnd)"),
    prob_brownian = md("*P*(BM)")
  ) %>% 
  as_latex

\D, Fritz and Purvis' *D* [@Fritz2010]; *P*(rnd), probability of random distribution of traits; *P*(BM), probability of traits evolving under Brownian motion (BM). Values of \D 0 or less indicate evolution according to BM; values near 1 indicate random trait distribution.

@@

pic_results %>%
  mutate_at(vars(estimate, std.error, statistic), ~round_t(., 2)) %>%
  mutate(p.value = format_pval(p.value)) %>%
  select(trait, n_contrasts, n_pos_con, statistic, p.value) %>%
  mutate(trait = factor(
    trait, 
    levels = c("stipe", "length", "width", "rhizome", 
               "sla", "pinna", "dissection"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Stipe length", "Frond length", "Frond width", "Rhizome diam.", 
                   "SLA", "Pinna no.", "Frond dissection"))) %>%
  arrange(trait) %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    n_contrasts = "Num. constrasts",
    n_pos_con = "Num. pos. constrasts",
    statistic = md("*t*"),
    p.value = md("*P*"),
  ) %>% 
  as_latex
```

All trait contrasts were made by subtracting epiphytic values from terrestrial values; thus, a positive contrast indicates greater values for terrestrial clades.

@@

correlated_evo_test %>%
  mutate(pval = format_pval(pval)) %>%
  mutate_if(is.numeric, ~round_t(., 1)) %>%
  mutate(trait = factor(
    trait, 
    levels = c("gemmae", "glands", "hairs","morphotype"))) %>%
  mutate(trait = lvls_revalue(
    trait,
    new_levels = c("Gemmae", "Glands", "Hairs", "Cordate morphotype"))) %>%
  arrange(trait) %>%
  gt(rowname_col = "trait") %>%
  cols_label(
    logL_indep = "LogL(independent)",
    logL_dep = "LogL(dependent)",
    likelihood_ratio = "Likelihood ratio",
    pval = md("*P*"),
  ) %>% 
  as_latex
```

LogL, log-likelihood. Each gametophyte trait was coded as a binary trait and tested for correlated evolution with growth habit (epiphytic vs. terrestrial growth). Traits with a significantly lower log likelihood for the dependent model are significantly correlated with growth habit.
